The Galactic Midplane Is Not a Plane: Implications for Dynamical Analysis with {\em Gaia} Data and Beyond

### 2019-04-26 paper/ms.tex
Fix line indentation throughout

Stellar actions, computed for stars from their six-dimensional phase space measurements and
an assumed Galactic potential, are used to label and distinguish orbits. In
principle, actions have the attractive quality of being invariant and thus
good labels of a star's orbit. 
%mm Of course,
    Such analyses are often used to interpret data from the {\em Gaia}
    mission. However, 
inaccurate measurements of the 
phase space position and uncertainties in the assumed Galactic potential
induce an error in the computed action. We show that, in addition to these
complications, a systematic bias in the phase space position induces a
phase-dependence in the actions, which we interpret as 
%mm an 
    a systematic
error. The induced error distribution is
non-Gaussian and bimodal, 
with neither of the modes peaking on
%mm  the true action value. [the error in a Gaussian peaks on zero] 
    the null value.
A variation in the vertical position midplane of $\sim15\,\pc$ for a thin disk orbit or
$\sim120\,\pc$ for a thick disk orbit is 
%mm all that is necessary 
      sufficient
to induce a $25\,\%$ systematic error in the vertical action $J_z$ computed under the assumption of an axisymmetric potential.
%mm [fragment?] orbits, respectively. 
Furthermore, we show that the local midplane varies by
$\sim100\,\pc$ at the Solar circle in Milky Way-mass cosmological zoom-in
simulations from the FIRE collaboration. Thus, current state-of-the-art action
calculations --- which assume the global and local midplanes are the same ---
include a systematic 
%mm offset in $z$.
   vertical offset variation. 
Variation in the local standard of rest
%mm will induce 
     induces
similar issues. The variation of the midplane must be taken into
account when performing dynamical modelling across
the large regions of the disk
accessible 
%mm with the {\em Gaia} mission. [moved first mention up to top of
%abstract, given that this is in the title.]
    to {\em Gaia}.

@@

Importantly, the location of Sgr~A\textsuperscript{*} may not be equivalent to the location of the ``dynamical Galactic center:'' the point in three-dimensional
space about which the stars in the solar neighborhood are orbiting. This assumption, although frequently made, has not yet been justified. It is tested in any construction of a  dynamical model where the distance from the Sun to
the Galactic center, $R_0$, is a free parameter.
For example, \citet{2015ApJ...803...80K} measured $R_0$ while
modelling the dynamics of the stream Palomar 5. Many other dynamical measurements
of $R_0$ have been made (\citealt{2016ARAA..54..529B} summarize many pre-{\em Gaia} results) but none have yet achieved a precision comparable to that of the distance to Sgr~A\textsuperscript{*}.
% \citep{1981gask.book.....M, 2011PASJ...63..867S, 2012MNRAS.427..274S,2013AstL...39...95B, 2013IAUS..289..444Z}.

>>>>

Stellar actions, computed for stars from their six-dimensional phase space
measurements and an assumed Galactic potential, are used to label and
distinguish orbits. In principle, actions have the attractive quality of being
invariant and thus good labels of a star's orbit. Such analyses are often used
to interpret data from the {\em Gaia} mission. However, inaccurate
measurements of the phase space position and uncertainties in the assumed
Galactic potential induce an error in the computed action. We show that, in
addition to these complications, a systematic bias in the phase space position
induces a phase-dependence in the actions, which we interpret as a systematic
error. The induced error distribution is non-Gaussian and bimodal, with
neither of the modes peaking on the null value. A variation in the vertical
position midplane of $\sim15\,\pc$ for a thin disk orbit or $\sim120\,\pc$ for
a thick disk orbit is sufficient to induce a $25\,\%$ systematic error in the
vertical action $J_z$ computed under the assumption of an axisymmetric
potential. Furthermore, we show that the local midplane varies by
$\sim100\,\pc$ at the Solar circle in Milky Way-mass cosmological zoom-in
simulations from the FIRE collaboration. Thus, current state-of-the-art action
calculations --- which assume the global and local midplanes are the same ---
include a systematic vertical offset variation. Variation in the local
standard of rest induces similar issues. The variation of the midplane must be
taken into account when performing dynamical modelling across the large
regions of the disk accessible to {\em Gaia}.

@@

Importantly, the location of Sgr~A\textsuperscript{*} may not be equivalent to
the location of the ``dynamical Galactic center:'' the point in
three-dimensional space about which the stars in the solar neighborhood are
orbiting. This assumption, although frequently made, has not yet been
justified. It is tested in any construction of a  dynamical model where the
distance from the Sun to the Galactic center, $R_0$, is a free parameter. For
example, \citet{2015ApJ...803...80K} measured $R_0$ while modelling the
dynamics of the stream Palomar 5. Many other dynamical measurements of $R_0$
have been made (\citealt{2016ARAA..54..529B} summarize many pre-{\em Gaia}
results) but none have yet achieved a precision comparable to that of the
distance to Sgr~A\textsuperscript{*}.
